Draft Article

### 2020-09-08 article.tex
Add hypothesis and new Baltic Incident analysis

If removing a quarter of the officers doesn't do lasting damage, then why not remove the full 200? The swift expulsion of all suspected intelligence officers from a country---and the inevitable removal of the country's own intelligence officers in retaliation---would deeply injure both their espionage capabilities indefinitely. This nuclear option is always available to both parties and yet, in modern history, it has never been exercised. We can reasonably infer that each nation's leadership is making an identical calculation: the benefits of continuing to practice espionage in my adversary's backyard outweigh the risks associated with allowing my adversary to do the same to me. The question remains how they can both be correct.

@@

% https://tex.stackexchange.com/questions/249480/creating-a-payoff-matrix-using-latex-tabular-environment
\begin{table}[ht]
\centering
\setlength{\extrarowheight}{2pt}
\small
\begin{tabular}{cc|c|c|}
  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{State 2}\\
  & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Tolerate Espionage}  & \multicolumn{1}{c}{Punish Espionage} \\\cline{3-4}
  \multirow{3}*{State 1}  & Tolerate Espionage & \makecell{~\\Both sides gain intelligence \\~} & \makecell{~\\ Only State 2 continues spying \\ ~} \\\cline{3-4}
  & Punish Espionage & \makecell{~\\ Only State 1 continues spying \\~} & \makecell{~\\ No intelligence either direction \\~} \\\cline{3-4}
\end{tabular}
\caption{Espionage consequences decision matrix}
\label{espionage-decision-matrix}
\end{table}

@@

To arrive at mutual cooperation, a security-seeking state needs to not only determine that its rival is a security seeker, it needs its rival to make that exact same judgment, because it is the adversary's response to securitization that ultimately makes the first state insecure. Defensive realism therefore introduces a counterintuitive conclusion: it is often logical for a security seeker to take actions that increase not only its own security, but its \emph{adversary's} security as well.\footcite[p.~7]{glaser_rational_2010} Such actions would make the adversary less likely to securitize, and also signal to the adversary that it is a security seeker hoping to reach a mutually agreeable outcome. If the adversary is indeed a security seeker, then, no longer concerned about its security, it will respond in kind and resolve the security dilemma for both of them. The problem for a defensive realist is that unilaterally taking steps to increase the adversary's security is obviously quite risky. The state could be wrong that its adversary is a security seeker---what Glaser calls a \enquote{greedy state}---in which case its optimistic overtures of cooperation will have exposed it to greater risk of attack. When states are unwilling to make those signaling measures because they fear compromising their own security, then a security dilemma results.

@@

States take the risk that espionage will reveal more than just a state's (hopefully benign) intent---it might also reveal a state's capabilities and strategy. For example, I interviewed senior Obama Administration official Jake Sullivan about the benefits of allowing foreign states to spy on the US. According to him, the United States would have no problem with adversaries being able to independently verify American intent because, as he sees it, American intent is benign. Capabilities, however, are more complicated. Some defense capabilities would probably be fine to have an adversary independently verify, and some the US might want to keep secret because their secrecy confers a strategic advantage. And under almost no circumstances would the US want to have its strategies and warfighting plans revealed, so every measure should be taken to ensure their secrecy.\footcite{sullivan_personal_2019} States would prefer to reveal only the \enquote{good} information, but they can't---the credibility of the information obtained depends on the fact that the espionage subverted its target's access controls. If you knew your target had let you find something, you wouldn't trust it.

@@

\subsection{Hypothesis}

Previous intelligence scholars have noted specific benefits of espionage that support the abstract theoretical justification I build here. Christopher Baker explains how espionage makes functional cooperation possible through treaties.\footcite[p.~6]{baker_tolerance_2004} Charles Glaser uses the 1972 Anti-Ballistic missile treaty as an example of precisely the type of signal that could lead states to resolve the security dilemma Julius Stone, a professor of international law, foresaw this exact role for espionage in 1962: \enquote{If you do not have a system of international inspection and if you can't get one (and I think it is quite likely that we can't) then the function which international inspection is supposed to serve still needs fulfilling. You may not be able to fulfill it to the optimum extent by reciprocally tolerated espionage, but you may be able to reduce the dangers.}\footcite[p.~41]{stone_legal_1962}

The best way to demonstrate that this theory has explanatory power for state decision-making is to find historical examples where states made decisions that showed they were willing to accept the cost of enemy espionage in order to practice espionage themselves. If the theory is correct, we would find evidence of policymakers justifying their decisions based on the value of mutual observation. Such decisions were made frequently during the Cold War, and with the explicit intent to increase the bilateral flow of information to create a more secure world environment. This article will present two such cases: aerial espionage and spy satellites.

Both planes and satellites played an under-explored role in the formulation of some of the Cold War's landmark diplomatic achievements. SALT I (the ABM Treaty) is widely considered to be the landmark diplomatic achievement of Nixon's \emph{d\'etente} policy, which set the stage for a broader easing of relations between the US and the USSR, and spy satellites were crucial to reducing the cost that each side would pay to sign it. Limiting ballistic missile defense systems was a relatively minor concession for both sides, but in the highly risk-averse context of the Cold War, the signal was significant.\footcite[p.~66]{glaser_rational_2010} Facing the omnipresent threat of nuclear annihilation, \emph{d\'etente} provided a rational strategic choice for both states---made possible by their reciprocal tolerance of espionage.

@@

% because it allows us to independently verify the intent and capabilities of other nations. Doing so actually reduces the likelihood that miscalculation will lead to open hostilities. As a result, both states allow it to happen, because they each benefit from a less-tense security environment. Stone aptly described this phenomenon as \enquote{reciprocally tolerated espionage.}

@@

\subsection{Aerial espionage}
The line between a \enquote{cold} and a \enquote{hot} war is supposed to be obvious. A superpower's own uniformed military personnel opening fire on the other superpower's military or territory would be \enquote{hot,} everything else would be \enquote{cold.} Broadly speaking, the Cold War is remembered as lacking direct military confrontation between the United States and Russia; though the two superpowers of the Cold War were open rivals, they supposedly competed for influence by funding, arming, or otherwise influencing proxies, rather than attacking each other directly. That impression is not exactly accurate. Not only did the US and USSR often come close to openly attacking each other, but a surprising number of times they actually did—--including one of the Cold War's most famous moments.

% citation needed on the definition of cold

Pilot Francis Gary Powers knew that U-2 overflights of the Soviet Union were incredibly risky. The Soviets were deeply unhappy that the US was violating their airspace in such a brazen fashion, but for years they had kept quiet about the U-2 missions, embarrassed that they were technologically incapable of intercepting the high-flying aircraft.\footcite[p.~59]{powers_operation_2004} When Soviet air defenses finally managed to shoot one down in 1960, Soviet leadership was equal parts triumphant and furious. At a planned summit in Paris only a few weeks later, Soviet Premier Khrushchev launched into a rant about the U-2 planes and stormed out on the first day, effectively ending the summit---and any hope that pressing issues like the fate of West Berlin would be resolved there. He also canceled a planned visit by US President Dwight Eisenhower to the USSR that was scheduled to take place the next month. A contemporary newsreel about the conference said that \enquote{in the course of two hours, Khrushchev brought US-Soviet relations to their lowest point since the end of World War II.}\footcite{universal_studios_summit_1960}

The shootdown of Gary Powers, in what would later come to be known as \enquote{The U-2 Incident}, clearly violated the only \enquote{rule} of the Cold War. Official Soviet air defenses shot down a US aircraft. All parties involved had cause to escalate militarily---the Soviets for the violation of their airspace, the US for the attack on their \enquote{weather craft}---but instead they chose to treat the U-2 missions as an intelligence operation, not a military action. The Soviets therefore applied the traditional consequences associated with espionage; Khrushchev engaged in some saber-rattling and denounced the violation of his airspace, but no material actions were taken, and the United States escaped the incident with few tangible consequences. The striking disparity between the diplomatic furor and the lack of material consequences can be explained by an apparent technicality: the U-2 was not a military aircraft, it was a spy plane.

No single attribute defines a spy plane. A military aircraft can forgo weapons but, by virtue of its pilot, its mission, its timing, or its branding still conduct what is unquestionably a military operation. Inversely, intelligence missions can be flown by uniformed Air Force pilots in re-purposed B-52s and somehow still be understood as conducting espionage. A spy plane is an ontological designation, one that reveals itself to the adversary through a series of social and physical signifiers that are not replicable across time. Germany in 1937 might not have reacted to a U-2 overflight the same way that the Soviet Union did in 1960. But even if the way different states handled espionage was perfectly consistent, there still wouldn't be a checklist to guarantee that a mission would be treated as espionage; it is not enough make a military mission look like an intelligence operation. The reality is both simpler and more frustrating---it has \emph{to be one}.

The U-2 incident was the most visible in a series of aerial intelligence operations that the United States undertook early in the Cold War. The US did not always employ civilian pilots nor it did always use obviously weaponless aircraft, yet it managed to run a shocking number of missions that violated Soviet airspace without significantly altering the relative level of tension between the two superpowers. On a surprisingly frequent basis, these missions would turn \enquote{hot} when Soviet air defenses shot them down. Not once, however, did these shootdowns lead to retaliation from the US, or significant pressure from the Soviets to stop the flights. Instead, each one was resolved as if the pilots who survived were simply spies, captured in the normal course of counterintelligence. This article seeks to answers how and why states ensure that certain operations reveal themselves as espionage; the decade of covert aerial espionage was one such successful operation.

\subsubsection{Military reconnaissance flights}
The tense battle between the US and USSR over surveillance of Soviet border installations was never formally acknowledged by either party---and neither were its combatants. Over the course of the Cold War, more than 200 American pilots were shot down while spying on the Soviet Union.\footcite{glenshaw_secret_2017} These missions were so highly classified that in many cases the families of the pilots never learned how they died. 126 of these American airmen remain unaccounted for today, casualties of missions whose existence both the US government and the Soviet Union collaborated for decades to deny.

Though they flew Air Force planes, these men were actually conducting one of the earliest forms of Cold War peacetime espionage. Their purpose was to gather SIGINT (signals intelligence) identifying the location of critical radar installations along the Soviet border. Converted bombers---called \enquote{ferrets} by the Air Force---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} According to the CIA, these missions began as early as 1947.\footcite[p.~4]{peterson_maybe_1993} Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were \enquote{overflights,} missions that deliberately violated Soviet airspace. The earliest definitive postwar overflight of the Soviet Union took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}

Without access to classified archives it is impossible to state for certain exactly how many of these flights were conducted. Even if one did have unlimited access, some of the necessary information was so highly classified that it is likely lost forever. Nonetheless, the civilian researcher is aided by the declassified research of military archivists before them and the works of two scholars in particular are highlighted here. First, Air Force Academy historian Dr. John Farquhar specializes in aerial reconnaissance and its diplomatic consequences. Many of the incidents analyzed in this subsection are taken from newspaper clippings that he compiled and analyzed. Second, former intelligence analyst and section chief Michael Peterson wrote an article for the CIA history magazine \emph{Cryptologic Quarterly} that analyzed the Soviet shootdowns of US reconnaissance aircraft in depth.

Peterson's article was written for a classified audience in 1993 and declassified in 2009, providing the public with a comprehensive (as far as we know) list of all American SIGINT flights shot down by the USSR (Table \ref{soviet-shootdowns}). Peterson uses two limiting criteria that perfectly suit the purposes of this article: the only flights he lists are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) explicitly US reconnaissance aircraft.\footcite[p.~4]{peterson_maybe_1993} Note that not all of these flights are necessarily overflights; many took place over international waters when they were shot down. In effect, the table catalogs the 13 times where the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5. In the original document, this table lists the first incident as having taken place over the Barents Sea, not the Baltic Sea. Because the description of the mission---including a map of its route in the same document---takes place entirely over the Baltic sea, I have concluded that this has to be a typographical error, and corrected it here.]{peterson_maybe_1993}

\begin{table}[ht]
\centering
\begin{tabular}{llr}
\textbf{Date}     & \textbf{\makecell[l]{U.S. Service \&\\ Aircraft Type}}   & \textbf{General Location} \\
8 April 1950      & USN PB4Y2 Privateer           & Baltic Sea                           \\
6 November 1951   & USN P2V Neptune               & Sea of Japan                         \\
13 June 1952      & USAF RB-29                    & Sea of Japan                         \\
7 October 1952    & USAF RB-29                    & East of Hokkaido/Kuril Is.           \\
29 July 1953      & USAF RB-50                    & Sea of Japan                         \\
4 September 1954  & USN P2V Neptune               & Sea of Japan                         \\
7 November 1954   & USAD RB-29                    & East of Hokkaido.Kuril Is.           \\
18 April 1955     & USAF RB-47                    & Off Kamchatka Peninsula              \\
10 September 1956 & USAF RB-50                    & Sea of Japan                         \\
2 September 1958  & USADF C-130                   & Soviet Armenia
\\
1 May 1960        & CIA U-2                       & Sverdolsk, USSR                      \\
1 July 1960       & USAF RB-47                    & Barents Sea                          \\
10 March 1964     & USAF RB-66                    & East Germany
\end{tabular}
\caption{Summary of Soviet Shootdowns, 1950-1964}
\label{soviet-shootdowns}
\end{table}

Regardless of how thorough declassified CIA information appears, one cannot rule out the possibility that there were other, even more secret missions which, for whatever reason, the United States government has chosen to conceal to this day. Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article are promising, and suggest that the information is relatively complete. First, the article was written for a classified audience as an overview of reconnaissance shootdowns during the 1950s, so it is likely intended to be comprehensive. Shootdowns are a sufficiently significant event such that it is unlikely the author would have just missed one. Additionally, when the article was declassified 16 years later, it appears to have been declassified without redactions or missing pages. Lastly, if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.

To comprehensively analyze the diplomatic effects of aerial espionage, I cross-referenced the shootdowns that the CIA says took place with Farquhar's analysis of their media and diplomatic impact. A shootdown is either present in both lists, or just one of the two---some incidents in the CIA list are missing from Farquhar's list because they did not have a diplomatic impact, and some incidents that Farquhar analyzes are not mentioned by Peterson because they were not SIGINT reconnaissance missions and likely (though not definitively) served little espionage function. Together, the full spectrum of aerial reconnaissance incidents demonstrates a clear pattern in which shootdowns that were interpreted as espionage were quickly resolved by both parties, while the shootdown of a flight unrelated to espionage produced swift, significant backlash.

The first of these shootdowns, on April 1950, brought the previously-hidden aerial reconnaissance program to the attention of the American people. An unarmed Navy Patrol plane, a PB4Y2 Privateer, was intercepted and downed by Soviet fighters over the Baltic sea. The Soviet ambassador to the US lodged a formal note of protest, incorrectly alleging that the US aircraft which had been shot down was a B-29 bomber.\footcite{kirk_ambassador_1950} A few days later, the American ambassador issued a formal response, in which he correctly identified the model of plane and claimed that at no point did it cross into any territorial waters. He demanded the Soviets conduct \enquote{a prompt and thorough investigation} and \enquote{see to it that those responsible for this action [were] promptly and severely punished.}\footcite{the_associated_press_text_1950} The Soviet Union recognized this for the toothless statement that it was, and rejected the demand in its entirety.\footcite{salisbury_kremlin_1950} No further action was taken.

Both sides had reasons to make this a larger political issue. The Soviet Union had fired first on a US Navy aircraft---which could not have been the aggressor because its only armament was a .45-cal pistol carried by one of the crewmen.\footcite[p.~7]{peterson_maybe_1993} Given that the plane was never found, all available evidence suggests that the flight likely did not violate the traditional 12-mile airspace boundary. Based on the last received position report, today the CIA claims that the plane was flying 20-25 miles off the coast of Latvia at the time it was shot down.\footcite[p.~7]{peterson_maybe_1993} At the time, it was theorized that a theoretical breakdown in the plane's navigation system could have caused it to veer off-course into Latvia, but given its last known position, it would have taken \enquote{a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.}\footcite{the_new_york_times_soviet_1950} Meanwhile, the United States knew that the missing plane probably hadn't ended up in Soviet airspace, but the Privateer in question \emph{had} been performing a reconnaissance mission of questionable legality when it was shot down. No traditional boundaries were violated, but a repurposed Navy privateer loaded with electronic reconnaissance equipment flying close to Latvia looked, to use a less technical term, shady. And still, based on the initial exchange of diplomatic protests, neither side was inclined to magnify the issue if they could avoid it.

>>>>

If removing a quarter of the officers doesn't do lasting damage, then why not remove the full 200? The swift expulsion of all suspected intelligence officers from a country---and the inevitable removal of the country's own intelligence officers in retaliation---would deeply injure both their espionage capabilities indefinitely.\footnote{Nothing on this scale has ever been attempted, but Operation Foot, the largest diplomatic expulsion in the Cold War, \enquote{was a blow from which the Soviet intelligence effort in the UK never recovered.} For more, see: \cite{hughes_giving_2006-1}} This nuclear option is always available to both parties and yet, in modern history, it has never been exercised. We can reasonably infer that each nation's leadership is making an identical calculation: the benefits of continuing to practice espionage in my adversary's backyard outweigh the risks associated with allowing my adversary to do the same to me. The question remains how they can both be correct.

@@

% https://tex.stackexchange.com/questions/249480/creating-a-payoff-matrix-using-latex-tabular-environment
\begin{table}[ht]
\centering
\setlength{\extrarowheight}{2pt}
\small
\begin{tabular}{cc|c|c|}
  & \multicolumn{1}{c}{} & \multicolumn{2}{c}{State 2}\\
  & \multicolumn{1}{c}{} & \multicolumn{1}{c}{Tolerate Espionage}  & \multicolumn{1}{c}{Punish Espionage} \\\cline{3-4}
  \multirow{3}*{State 1}  & Tolerate Espionage & \makecell{~\\Both sides gain intelligence \\~} & \makecell{~\\ Only State 2 continues spying \\ ~} \\\cline{3-4}
  & Punish Espionage & \makecell{~\\ Only State 1 continues spying \\~} & \makecell{~\\ Limited intelligence either direction \\~} \\\cline{3-4}
\end{tabular}
\caption{Espionage consequences decision matrix}
\label{espionage-decision-matrix}
\end{table}

@@

To arrive at mutual cooperation, a security-seeking state needs to not only determine that its rival is a security seeker, it needs its rival to make that exact same judgment, because it is the adversary's response to securitization that ultimately makes the first state insecure. Defensive realism therefore introduces a counterintuitive conclusion: it is often logical for a security seeker to take actions that increase not only its own security, but its \emph{adversary's} security as well.\footcite[p.~7]{glaser_rational_2010} If the adversary is indeed a security seeker, then, no longer concerned about its security, it will respond by not building up arms, and resolve the security dilemma for both of them. The problem for a defensive realist is that unilaterally taking steps to increase the adversary's security is obviously quite risky. The state could be wrong that its adversary is a security seeker---what Glaser calls a \enquote{greedy state}---in which case its optimistic overtures of cooperation will have exposed it to greater risk of attack. When states are unwilling to make those signaling measures because they fear compromising their own security, then a security dilemma results.

@@

States take the risk that espionage will reveal more than just a state's (hopefully benign) intent---it might also reveal a state's capabilities and strategy. For example, I interviewed senior Obama Administration official Jake Sullivan about the benefits of allowing foreign states to spy on the US. According to him, the United States would have no problem with adversaries being able to independently verify American intent because, as he sees it, American intent is benign. Capabilities, however, are more complicated. Some defense capabilities would probably be fine to have an adversary independently verify, and some the US might want to keep secret because their secrecy confers a strategic advantage. And under almost no circumstances would the US want to have its strategies and warfighting plans revealed, so every measure should be taken to ensure their secrecy.\footcite{sullivan_personal_2019} States might prefer to reveal only the \enquote{good} information, but they can't. Espionage is by definition illicit---the fact the information was obtained without the target's permission lends it credibility.

@@

\subsection{Hypothesis}
In the following section, I will use empirical case studies to demonstrate how associating an operation with espionage alters its diplomatic response. In order to guide that research, it is important to establish a series of clear, testable hypotheses. I first present a null hypothesis:

\begin{quote}
$H_0 =$ Associating a state action with the traditional practice of espionage will not meaningfully alter the diplomatic response.
\end{quote}

The null hypothesis posits that there is insufficient evidence linking the perception that an act \enquote{is} espionage to the response that the affected state ultimately delivers. If this hypothesis holds, espionage may have some symbolic importance, but it isn't one that can be meaningfully tied to a diplomatic response. By contrast, here is the alternative hypothesis, the one that would prove true if the theory holds:

\begin{quote}
$H_1 =$ Associating a state action with espionage will result in a reduced diplomatic response when compared with equivalent actions that do not share that association.
\end{quote}

The argument presented by this article risks a tautology: perhaps we associate certain actions with espionage \emph{because} they received a reduced diplomatic response. For this reason, the relationship in $H_1$ is explicitly framed as causal. Unfortunately, defining the relationship that way makes it difficult to prove directly; no two state actions take place under equivalent circumstances, so no natural experiment could demonstrate that equivalent actions, perceived differently, received different responses. I therefore propose a corollary hypothesis to aid us in proving the alternative hypothesis:

\begin{quote}
$H_2 =$ States purposefully pursue measures to signal that their actions constitute espionage, because they apply reduced diplomatic responses to espionage and hope that their adversary will do the same.
\end{quote}

Proving $H_2$ is significantly easier than proving $H_1$. State actions---and the reasons policymakers provided for pursuing those actions---are often documented in the historical record. Careful research could demonstrate that \enquote{espionage} is an ontologically significant concept, not just for academics, but for the very people who engage in it. If $H_2$ holds, one would find evidence of policymakers being willing to accept the cost of enemy espionage in order to practice espionage themselves based on the value of mutual observation. Furthermore, when $H_2$ is proven to be true bidirectionally, it implies $H_1$. Two states that are both maintaining reduced responses to espionage (while pursuing espionage themselves) create a relationship in which $H_1$ holds: state actions successfully associated with espionage will receive a reduced response.

% Previous intelligence scholars have noted specific benefits of espionage that support the abstract theoretical justification I build here. Christopher Baker explains how espionage makes functional cooperation possible through treaties.\footcite[p.~6]{baker_tolerance_2004} Charles Glaser uses the 1972 Anti-Ballistic missile treaty as an example of precisely the type of signal that could lead states to resolve the security dilemma Julius Stone, a professor of international law, foresaw this exact role for espionage in 1962: \enquote{If you do not have a system of international inspection and if you can't get one (and I think it is quite likely that we can't) then the function which international inspection is supposed to serve still needs fulfilling. You may not be able to fulfill it to the optimum extent by reciprocally tolerated espionage, but you may be able to reduce the dangers.}\footcite[p.~41]{stone_legal_1962}

The following section will demonstrate that both the United States and the USSR were forced to contend with the benefits and risks of a mutually permissive espionage regime during the Cold War. When faced with opportunities to either shut down the espionage apparatus or expand it, they both chose to expand it. Policymakers on either side of the Iron Curtain repeatedly decided that preserving the bilateral flow of information ensured a more secure world environment. This article will present two such cases: aerial espionage and spy satellites.

% Both planes and satellites played an under-explored role in the formulation of some of the Cold War's landmark diplomatic achievements. SALT I (the ABM Treaty) is widely considered to be the landmark diplomatic achievement of Nixon's \emph{d\'etente} policy, which set the stage for a broader easing of relations between the US and the USSR, and spy satellites were crucial to reducing the cost that each side would pay to sign it. Limiting ballistic missile defense systems was a relatively minor concession for both sides, but in the highly risk-averse context of the Cold War, the signal was significant.\footcite[p.~66]{glaser_rational_2010} Facing the omnipresent threat of nuclear annihilation, \emph{d\'etente} provided a rational strategic choice for both states---made possible by their reciprocal tolerance of espionage.

@@

@@

\subsection{Aerial espionage}

Just a few weeks before the \enquote{Four Powers Summit} in Summer 1960, Soviet Premeir Nikita Khrushchev announced that the USSR had shot down an American plane.\footcite[p.~112]{powers_operation_2004} The CIA knew immediately that it had lost a U-2, an advanced spy plane that had bedeviled the Soviets for years, proving beyond the reach of their surface-to-air missiles.\footcite{orlov_u-2_2007} Though clearly the technological gap had now closed, the CIA was still confident that its cover story would hold, thanks to a morbid fact of the U-2's design: in the event of a shootdown, it should have been impossible for the pilot to survive.\footcite[p.~35]{lindgren_trust_2000} When the United States put out at a statement that the plane was part of a \enquote{weather reconnaissance} mission gone awry, Khrushchev sprung his trap and revealed that the U-2 pilot, Gary Powers, had been taken alive.

In a few days, the previously pleasant trend in US-Soviet relations completely reversed. President Dwight D. Eisenhower admitted that the Gary Powers had been flying a spy plane, declaring in a chagrined press conference that espionage was ``a distasteful but vital necessity'' to prevent another attack like Pearl Harbor.\footcite{eisenhower_news_1960} At the planned summit, Khrushchev launched into a tirade about U-2 planes and walked out on the first day, effectively ending the summit. He also canceled a planned visit by US President Dwight Eisenhower to the USSR that was scheduled to take place the next month. A contemporary newsreel about the conference said that \enquote{in the course of two hours, Khrushchev brought US-Soviet relations to their lowest point since the end of World War II.}\footcite{universal_studios_summit_1960}

In the traditional telling, the U-2 Incident is often cited as a major failure of the Eisenhower administration, which significantly worsened US-Soviet relations at time when lasting peace appeared possible. Only in the under-studied context of Cold War reconnaissance flights does it become clear that the U-2 Incident actually shows both the US and USSR exhibiting restraint and sound judgment. Over a period of about 15 years, they each allowed provocative aerial intelligence operations to take place without military escalation. These operations were permitted because, consistent with $H_2$, the United States took great pains to signal that its various actions were within the bounds of espionage. The Soviet Union responded by honoring that signal and not escalating the situation beyond the immediately necessary measures to counteract it. The U-2 Incident should be understood as a uniquely embarrassing but otherwise typical failure of an intelligence operation---the maximal case for the minimal consequences typically applied to espionage.

% The shootdown of Gary Powers clearly violated the only \enquote{rule} of the Cold War. Official Soviet air defenses shot down a US aircraft. All parties involved had cause to escalate militarily---the Soviets for the violation of their airspace, the US for the attack on their \enquote{weather craft}---but instead they chose to treat the U-2 missions as an intelligence operation, not a military action. The Soviets therefore applied the traditional consequences associated with espionage; Khrushchev engaged in some saber-rattling and denounced the violation of his airspace, but no material actions were taken, and the United States escaped the incident with few tangible consequences. The striking disparity between the diplomatic furor and the lack of material consequences can be explained by an apparent technicality: the U-2 was not a military aircraft, it was a spy plane.

% No single attribute defines a spy plane. A military aircraft can forgo weapons but, by virtue of its pilot, its mission, its timing, or its branding still conduct what is unquestionably a military operation. Inversely, intelligence missions can be flown by uniformed Air Force pilots in re-purposed B-52s and somehow still be understood as espionage. Espionage is an ontological designation, one that reveals itself to the adversary through a series of social and physical signifiers that are not replicable across time. Germany in 1937 would likely have reacted to a U-2 overflight differently than the Soviet Union did in 1960. But even if the way different states handled espionage was perfectly consistent, there still wouldn't be a checklist to guarantee that a mission would be treated as espionage. It is not enough make a military mission look like an intelligence operation; the reality is both simpler and more frustrating---it has \emph{to be one}.

\subsubsection{Military reconnaissance flights}
Beginning in 1947, the US began a campaign of reconnaissance flights that routinely violated Soviet airspace. Converted Air Force bombers---called \enquote{ferrets}---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} The pilots who flew these missions were conducting one of the earliest forms of Cold War peacetime espionage: gathering SIGINT (signals intelligence) that identified the location of critical radar installations along the Soviet border. Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were \enquote{overflights,} missions that deliberately violated Soviet airspace.

The tense battle over Soviet border surveillance was never formally acknowledged by either state---and neither were its combatants. Over the course of the Cold War, more than 200 American pilots were shot down while spying on the Soviet Union.\footcite{glenshaw_secret_2017} These missions were so highly classified that in many cases the families of the pilots never learned how they died. 126 of these American airmen remain unaccounted for today, casualties of missions whose existence both the US government and the Soviet Union collaborated for decades to deny. Not once did these shootdowns lead to retaliation from the US, or significant pressure from the Soviets to stop the flights. Instead, the cases resolved as if the pilots were spies, captured or killed in the course of counterintelligence, whereabouts undisclosed.

% The earliest definitive postwar overflight of the Soviet Union took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}

Without access to classified archives it is impossible to state for certain exactly how many of these flights were conducted.\footnote{Some of the necessary information was so highly classified that it is likely lost forever.} Nonetheless, the civilian researcher is aided by the declassified research of military and intelligence archivists. Writing for the internal CIA publication \emph{Cryptologic Quarterly}, Michael Peterson compiled a now-declassified list of American SIGINT flights shot down by the USSR (Table \ref{soviet-shootdowns}).\footnote{Michael Peterson is a former intelligence analyst and section chief for the CIA.} Peterson uses two limiting criteria that perfectly suit the purposes of this article: the only flights he lists are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) involved US aircraft exclusively performing reconnaissance.\footcite[p.~4. Regardless of how thorough declassified CIA information appears, one cannot rule out the possibility that there were other, even more secret missions which, for whatever reason, the United States government has chosen to conceal to this day. Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article suggest that the information is relatively complete. First, the article was written for a classified audience, likely given that designation so that it could be comprehensive, and declassified after the fact. Additionally, when the article was declassified 16 years later, it appears to have been declassified without redactions or missing pages. Lastly, if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.]{peterson_maybe_1993} In effect, the table catalogs the 13 times where the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5. In the original document, this table lists the first incident as having taken place over the Barents Sea, not the Baltic Sea. Because the description of the mission---including a map of its route in the same document---takes place entirely over the Baltic sea, I have concluded that this must be a typographical error, and corrected it here.]{peterson_maybe_1993}

\begin{table}[ht]
\centering
\begin{tabular}{llr}
\textbf{Date}     & \textbf{\makecell[l]{U.S. Service \&\\ Aircraft Type}}   & \textbf{General Location} \\
8 April 1950      & USN PB4Y2 Privateer           & Baltic Sea                           \\
6 November 1951   & USN P2V Neptune               & Sea of Japan                         \\
13 June 1952      & USAF RB-29                    & Sea of Japan                         \\
7 October 1952    & USAF RB-29                    & East of Hokkaido/Kuril Is.           \\
29 July 1953      & USAF RB-50                    & Sea of Japan                         \\
4 September 1954  & USN P2V Neptune               & Sea of Japan                         \\
7 November 1954   & USAD RB-29                    & East of Hokkaido. Kuril Is.           \\
18 April 1955     & USAF RB-47                    & Off Kamchatka Peninsula              \\
10 September 1956 & USAF RB-50                    & Sea of Japan                         \\
2 September 1958  & USADF C-130                   & Soviet Armenia
\\
1 May 1960        & CIA U-2                       & Sverdolsk, USSR                      \\
1 July 1960       & USAF RB-47                    & Barents Sea                          \\
10 March 1964     & USAF RB-66                    & East Germany
\end{tabular}
\caption{Summary of Soviet Shootdowns, 1950-1964}
\label{soviet-shootdowns}
\end{table}

Peterson's article represents the CIA's internal accounting for SIGNIT shootdowns, but it does not tell which missions were perceived as significant by the Soviet Union. For that, I cross-referenced the CIA's list with Air Force historian Dr. John Farquhar's analysis of Cold War shootdowns that had a diplomatic impact.\footcite[Dr. John T Farquhar is a retired Lieutenant Colonel in the United States Air Force, and currently an Associate Professor of Military \& Strategic Studies at the US Air Force Academy.]{farquhar_aerial_2015} The combination of these two lists is instructive---some incidents in the CIA list are missing from Farquhar's list because they did not have a diplomatic impact, and some incidents that Farquhar analyzes are not mentioned by Peterson because they were not SIGINT reconnaissance missions. Taken together, the full spectrum of aerial reconnaissance incidents demonstrates a clear pattern: shootdowns that were interpreted as espionage were quickly resolved by both parties, while the shootdown of a flight unrelated to espionage produced swift, significant backlash.

% First, many of the incidents analyzed in this subsection are taken from clippings compiled by Air Force Academy historian Dr. John Farquhar, who specializes in aerial reconnaissance and its diplomatic consequences. Second,

The first of these shootdowns brought the aerial reconnaissance program to the attention of the American people on April 8, 1950.\footnote{Note that the flights analyzed here were not necessary overflights; many took place over international waters or dispute waters when they were shot down.} An unarmed Navy patrol plane was downed by Soviet fighters over the Baltic sea. The Soviet ambassador to the US lodged a formal note of protest.\footcite{kirk_ambassador_1950} A few days later, the American ambassador issued a formal response, claiming that at no point did the plane cross into any territorial waters. He demanded the Soviets conduct \enquote{a prompt and thorough investigation} and \enquote{see to it that those responsible for this action [were] promptly and severely punished.}\footcite{the_associated_press_text_1950} The Soviet Union recognized this for the toothless statement that it was, and rejected the demand in its entirety.\footcite{salisbury_kremlin_1950} No further action was taken.

Consistent with $H_2$, the United States took steps to code this flight as espionage, which was successful not just in the eyes of the Soviet Union, but to the American press as well. \emph{Washington Post} columnist Walter Lippmann wrote that \enquote{the known facts indicate that Soviet intelligence \textelp{} believed it was carrying important electronic equipment and that orders were given to the Soviet fighter command to intercept it.}\footcite{lippmann_baltic_1950} \enquote{Baltic Plane Mystery,} a \emph{Post} article written a few days later, was almost sympathetic to the Soviet commanders. \enquote{Electronics make the old delimitations for border coastal flights ridiculous. A plane flying on a course perfectly legal by standards accepted today might still be engaged in reconnaissance of the first importance that an unfriendly power would try to frustrate.}\footcite{childs_baltic_1950}

% This is true, both in terms of the mission itself and the technological environment, but he went even further. \enquote{The only sound and safe assumption is that the Russians have a thorough and far-reaching espionage system. And at the same time we must hope that our system, and particularly on the side of counterespionage, is effective.} Even the press, though outraged at the loss of American life, apparently took it as a given that this kind of espionage is necessary---and expected from both sides.
